Accesible, realistic genome simulation with selection using \stdpopsim

### 2025-03-11 analysis2_manuscript.tex
data availability; authors con't

\newcommand{\tskit}{\texttt{tskit}\xspace}

@@

\author[2]{Maria Izabel A. Cavassim}
% \author[5,7]{Jeff Adrion}
% \author[5]{Saurabh Belsare}
% \author[8]{Arjun Biddanda}
% \author[5]{Victoria Caudill}
% \author[9]{Jean Cury}
% \author[10]{Ignacio Echevarria}
% \author[11]{Benjamin C. Haller}
% \author[12,13]{Ahmed R. Hasan}
\author[14,15]{Xin Huang}
\author[5]{Kevin Korfmann}
% \author[16]{Leonardo Nicola Martin Iasi}
% \author[17]{Ekaterina Noskova}
% \author[18]{Jana Obšteter}
% \author[19]{Vitor Antonio Corrêa Pavinato}
% \author[20,21]{Alice Pearson}
% \author[22,23]{David Peede}
% \author[24]{Manolo F. Perez}
\author[5]{Chris C. R. Smith}
% \author[25]{Jeffrey P. Spence}
% \author[5]{Anastasia Teterina}

\author[5]{Scott T. Small}
\author[5]{Clara T. Rehmann}
% \author[26]{Per Unneberg}
% \author[27]{Juan Manuel Vazquez}
% \author[28]{Ryan K. Waples}
% \author[29]{Anthony Wilder Wohns}
% \author[30]{Yan Wong}
% \author[31]{Franz Baumdicker}
% \author[32]{Reed A. Cartwright}
% \author[33]{Gregor Gorjanc}
\author[4]{Kirk E. Lohmueller}
\author[34]{Ryan N. Gutenkunst}
\author[30]{Jerome Kelleher}
\author[35]{Aaron P. Ragsdale}

@@

% \affil[10]{\small{School of Life Sciences, University of Glasgow, Glasgow, UK}}
% \affil[11]{\small{Department of Computational Biology, Cornell University, Ithaca NY, USA}}

@@

% \affil[28]{\small{Department of Biostatistics, University of Washington, Seattle WA, USA}}

@@

% \affil[33]{\small{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, Edinburgh EH25 9RG, UK}}
\affil[34]{\small{Department of Molecular and Cellular Biology, University of Arizona, Tucson AZ 85721, USA}}

@@

    % implementing selection
    \stdpopsim previously provided a wide range of models of neutral
    demographic history from roughly two dozen species \citep{lauterbur2023expanding}.
    This was accomplished through the use of a standardized interface
    and data structure that allows for the curated addition of new
    species and models, and can use either
    \texttt{msprime} \citep{Baumdicker2022} or \texttt{SLiM} \citep{haller2019slim}
    as the backend engine for efficient population genetic simulation.
    To extend \stdpopsim's functionality to include models of selection, 
    we introduce two new classes
    to the \stdpopsim API: the \texttt{DFE} class and the \texttt{Annotation} class.
    The \texttt{DFE} class provides an interface for specifying
    a distribution of the fitness effects of new mutations.
    Each \texttt{DFE} object specifies a list of mutation types and
    their proportions, where 
    each mutation type is associated with a dominance coefficient
    (currently limited to values between 0 and 1, inclusive),
    the distribution of selection coefficients (e.g., fixed-value, normally-distributed, gamma-distributed),
    and the parameters of this distribution.
    Each \texttt{Annotation} object represents a collection of functional
    genomic elements (e.g., exons, introns, intergenic regions, etc.)
    whose coordinates are specified using a GFF3 file.      
    More concretely, consider implementing a model of selection consistent with the
    DFE inferred by Kim et al. (\citeyear{kim2017inference}) for exons in the human genome.
    The associated \texttt{DFE} object is defined with two mutation types,
    neutral and deleterious, corresponding to synonymous and non-synonymous sites.
    with their proportions set to follow their expected prevalence in exons. 
    The neutral mutation type is associated with a fixed selection coefficient (0),
    and the deleterious mutation type is associated with a dominance coefficient of $0.5$ and
    a gamma-distributed selection coefficient
    with mean and shape as inferred by Kim et al. (\citeyear{kim2017inference}).
    \texttt{DFE} objects can also model a relationship between the selection and dominance coefficients \citep{huber2018gene}
    by defining ranges of selection coefficients and associating each range with a different dominance coefficient (see below).

@@

    From a user's perspective, a model of selection is specified by pairing a \texttt{DFE} object
    with a collection of genomic segments defined using an \texttt{Annotation} object.
    Multiple such pairings can be defined in one simulation
    to provide a rich model of how selection may vary along a chromosome.
    For example, coding sequences may be associated with one DFE and non-coding sequences in exons with another DFE.
    Simulations with selection are then implemented by specifying a species, 
    a portion of the genome to simulate,
    a genetic map if available, a model of demography, and pairings of DFEs with genomic segments
    using either the python API or the command-line interface (Figure \ref{fig:schematic}B-C).
    Simulations with selection are generated
    using the \texttt{SLiM} simulation engine, which is perhaps
    the most flexible simulation engine for modeling selection available to date
    (\texttt{msprime}'s selection models are relatively limited and thus not currently used by \stdpopsim).

@@

    A challenge when comparing different inferred and simulated distributions of the selection coefficient ($s$)
    is differing conventions in the literature regarding its definition.
    The simulator \texttt{SLiM} and the inference tool \grapes define the selection coefficient $s$ such that a homozygote has fitness $1+s$.
    On the other hand, \dadicli and \polydfe define the selection coefficient such that a homozygote has fitness $1+ 2s$.
     %https://academic.oup.com/genetics/article/207/3/1103/5930676
    Moreover, inferred distributions of selection coefficients are typically scaled by an inferred ancestral
    effective population size $N_e$, and different methods assume different scaling.
    \dadicli infers the distribution of $2 N_e s$ while \polydfe and \grapes infer the distribution of $4 N_e s$.
    %https://academic.oup.com/genetics/article/207/3/1103/5930676, https://academic.oup.com/genetics/article/216/2/559/6066183
    For comparison, here we have normalized all inference results to the convention of \texttt{SLiM} (see \textbf{Methods}).
    % Add citation to fastDFE as the successor to polyDFE: https://academic.oup.com/mbe/article/41/5/msae070/7641109 ?

@@

    \subsection*{Whole genome simulations}
    In this study we performed simulations of the human and vaquita porpoise genomes
    using the \stdpopsim{} framework.
    All simulations were executed using the SLiM engine with a burn-in of $2N$ generations
    (where $N$ is the initial population size), and a scaling factor of 2 (to reduce running time).
    Each simulation was implemented using a relatively short python script, similar to the one shown in Figure 1B
    (see \textbf{Data availablity} below).
    \igcomment{We should add a link to the analysis2 repo in data/code availablity section}
    %https://github.com/popsim-consortium/analysis2/tree/main
    % DRS: Maybe we should say something about how easy it was to do these simulations, and maybe even show
    % some example code somewhere to prove our point? Could help emphasize the utility of this resource.
    % IG: added a note about this above.

    \subsubsection*{Simulations of complete human genomes}
    Simulation of human genomes used a genetic map from the HapMap Project
    \citep{international2007second} (\stdpopsim label \texttt{HapMapII\_GRCh38}) and
    the out of Africa (OOA) demographic model with archaic admixture from
    \cite{ragsdale2019models} (\stdpopsim label \texttt{OutOfAfricaArchaicAdmixture\_5R19}).
    Simulations with selection additionally applied a DFE model with gamma-distributed selection coefficients inferred by
    \cite{kim2017inference} (\stdpopsim label \texttt{Gamma\_K17}) to exons annotated by the HAVANA group release 104
    (\cite{ensembl2018}; \stdpopsim label \texttt{ensembl\_havana\_104\_exons}).
    Each simulation generate 100 diploid genomes from each of the three extant human populations (YRI, CEU, and CHB),
    with each genome consisting of 22 autosomes.
    Using these settings, we generated three replicate datasets without selection
    and three replicate datasets with selection.
    Three additional datasets, used in Figure \ref{fig:homsap-dfe.constant},
    were simulated with the same model of selection and a simple demographic model with one population with a constant size of $10,000$.
    Each of these simulations generated 100 diploid genomes.

@@

    % We used three software packages to infer the distribution of fitness effects (DFE) from genetic data:
    % \dadicli \citep{Huang2023,kim2017inference}, \polydfe \citep{tataru2020polydfe}, and \grapes \citep{galtier2016adaptive}.
    % Each software package was run using the same set of simulations.
    % For these analyses we used the exon annotations from the human and vaquita porpoise genomes respectively.
    % A key distinction between our analyses and those done in practice
    % is that we have complete knowledge of which sites are under selection in the simulation,
    % thus our inferences is a best case scenario. Empirical inference of the DFE is a much
    % harder problem as some mutations assumed to be free of selection
    % may in reality be under selection. As above, all methods were run using default settings.
    \igcomment{I edited the description below written by Silas. Please check to make sure it's still accurate.}
    As described above, each DFE-inference method scales selection coefficients slightly differently.
    We thus rescaled each program's output to obtain a selection coefficient ($s$) consistent with the one used by SLiM.
    The output of \grapes  was scaled by a factor of  $\mu / \theta$,
    and the output of \polydfe was saled by $2 \mu / \theta$, 
    where $\mu$ is the assumed mutation rate and 
    $\theta$ is the population-scaled mutation rate estimated by each method ($\theta = 4N_{e}\mu$).
    \igcomment{Do we really multiple $\alpha$ by $\nu$ in rescaling for \dadicli?}
    For \dadicli, we report $\alpha * \nu / (\theta / (4 \mu L p))$, where
    $\alpha$ and $\nu$ are the inferred shape and rate parameters of the \dadicli's inferred gamma distribution,
    $L$ is the total number of sites available for the analysis,
    $p$ is the proportion of sites under selection.

@@

    To test methods for detecting selective sweeps, we simulated 100 evenly distributed 5cM
    segments of human chromosome 1.
    As recombination rate varies across the genome, the size of the region in base pairs simulated varied, however the recombination distance,
    which is critical for the linked effects of a sweep, was held constant.
    Each segment was simulated in each of the following three settings:
    (1) neutral, (2) background selection (BGS), and (3) hard sweep with BGS.
    In all three settings, we applied a genetic map from the HapMap Project
    \citep{international2007second} (\stdpopsim label \texttt{HapMapII\_GRCh38}) and
    the three-population out of Africa (OOA) demographic model from
    \cite{gutenkunst2009inferring} (\stdpopsim label \texttt{OutOfAfrica\_3G09}).
    In the two settings with BGS, we applied the DFE model with gamma-distributed selection coefficients inferred by
    \cite{kim2017inference} (\stdpopsim label \texttt{Gamma\_K17}) to exons annotated by the HAVANA group release 104
    (\cite{ensembl2018}; \stdpopsim label \texttt{ensembl\_havana\_104\_exons}).
    Hard sweeps were modeled by generating a single beneficial mutation 
    with selection coefficient of $s = 0.03$ in the center of the 5cM genomic segment
    and discarding simulations in which the beneficial mutation did not reach (or exceed) a terminal frequency of $95\%$.
    For each of the 100 genomic segments of chromosome 1, we generated 1,000 replicate datasets under each of the three settings.
    As with the whole-genome simulations (see above), we used the SLiM engine with a burn-in of $2N$ generations
    (where $N$ is the initial population size), and a scaling factor of 2.
    Each simulation generated 10 diploid genomic segments for each of the two sampled populations: YRI and CEU.
    The sequences simulated with a hard sweep and BGS were used for sweep detection,
    and sequences simulated under the other two settings were used to calibrate the thresholds of the test statistics
    used by the three sweep-detection methods (see below).

@@

\section*{Data availability}\label{data_availability}

>>>>

\newcommand{\tskit}{\texttt{tskit}\xspace}
\newcommand{\slim}{\texttt{SLiM}\xspace}
\newcommand{\msprime}{\texttt{msprime}\xspace}

@@

\author[2]{Maria Izabel A. Cavassim}
% \author[5,7]{Jeff Adrion}
% \author[5]{Saurabh Belsare}
% \author[8]{Arjun Biddanda}
% \author[5]{Victoria Caudill}
% \author[9]{Jean Cury}
\author[10]{Peter D. Fields}
\author[11]{Benjamin C. Haller}
\author[14,15]{Xin Huang}
\author[30]{Ben Jeffrey}
\author[5]{Kevin Korfmann}
\author[5]{Jiseon Min}
% \author[16]{Leonardo Nicola Martin Iasi}
% \author[17]{Ekaterina Noskova}
% \author[18]{Jana Obšteter}
% \author[19]{Vitor Antonio Corrêa Pavinato}
% \author[20,21]{Alice Pearson}
% \author[22,23]{David Peede}
% \author[24]{Manolo F. Perez}
\author[5]{Chris C. R. Smith}
% \author[25]{Jeffrey P. Spence}
% \author[5]{Anastasia Teterina}

\author[5]{Scott T. Small}
\author[5]{Clara T. Rehmann}
% \author[26]{Per Unneberg}
% \author[27]{Juan Manuel Vazquez}
% \author[28]{Ryan K. Waples}
% \author[29]{Anthony Wilder Wohns}
\author[30]{Yan Wong}
% \author[31]{Franz Baumdicker}
%\author[32]{Reed A. Cartwright}
\author[33]{Gregor Gorjanc}
\author[28]{Christian D. Huber}

\author[34]{Ryan N. Gutenkunst}
\author[30]{Jerome Kelleher}
\author[4]{Kirk E. Lohmueller}
\author[35]{Aaron P. Ragsdale}

@@

\affil[10]{\small{The Jackson Laboratory, Bar Harbor, ME, 04609 USA}}
\affil[11]{\small{Department of Computational Biology, Cornell University, Ithaca NY, USA}}

@@

\affil[28]{\small{Department of Biology, Pennsylvania State University, University Park, PA 16802, USA}}

@@

\affil[33]{\small{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, Edinburgh EH25 9RG, UK}}
\affil[34]{\small{Department of Molecular and Cellular Biology, University of Arizona, Tucson, AZ 85721, USA}}

@@

    % implementing selection
    \stdpopsim previously provided a wide range of models of neutral
    demographic history from roughly two dozen species \citep{lauterbur2023expanding}.
    This was accomplished through the use of a standardized interface
    and data structure that allows for the curated addition of new
    species and models, and can use either
    \msprime \citep{Baumdicker2022} or \slim \citep{haller2019slim}
    as the backend engine for efficient population genetic simulation.
    To extend \stdpopsim's functionality to include models of selection, 
    we introduce two new classes
    to the \stdpopsim API: the \texttt{DFE} class and the \texttt{Annotation} class.
    The \texttt{DFE} class provides an interface for specifying
    a distribution of the fitness effects of new mutations.
    Each \texttt{DFE} object specifies a list of mutation types and
    their proportions, where 
    each mutation type is associated with a dominance coefficient
    (currently limited to values between 0 and 1, inclusive),
    the distribution of selection coefficients (e.g., fixed-value, normally-distributed, gamma-distributed),
    and the parameters of this distribution.
    Each \texttt{Annotation} object represents a collection of functional
    genomic elements (e.g., exons, introns, intergenic regions, etc.)
    whose coordinates are specified using a GFF3 file.      
    More concretely, consider implementing a model of selection consistent with the
    DFE inferred by Kim et al. (\citeyear{kim2017inference}) for exons in the human genome.
    The associated \texttt{DFE} object is defined with two mutation types,
    neutral and deleterious, corresponding to synonymous and non-synonymous sites.
    with their proportions set to follow their expected prevalence in exons. 
    The neutral mutation type is associated with a fixed selection coefficient (0),
    and the deleterious mutation type is associated with a dominance coefficient of $0.5$ and
    a gamma-distributed selection coefficient
    with mean and shape as inferred by Kim et al. (\citeyear{kim2017inference}).
    \texttt{DFE} objects can also model a relationship between the selection and dominance coefficients \citep{huber2018gene}
    by defining ranges of selection coefficients and associating each range with a different dominance coefficient (see below).

@@

    From a user's perspective, a model of selection is specified by pairing a \texttt{DFE} object
    with a collection of genomic segments defined using an \texttt{Annotation} object.
    Multiple such pairings can be defined in one simulation
    to provide a rich model of how selection may vary along a chromosome.
    For example, coding sequences may be associated with one DFE and non-coding sequences in exons with another DFE.
    Simulations with selection are then implemented by specifying a species, 
    a portion of the genome to simulate,
    a genetic map if available, a model of demography, and pairings of DFEs with genomic segments
    using either the python API or the command-line interface (Figure \ref{fig:schematic}B-C).
    Simulations with selection are generated
    using the \slim simulation engine, which is perhaps
    the most flexible simulation engine for modeling selection available to date
    (\msprime's selection models are relatively limited and thus not currently used by \stdpopsim).

@@

    A challenge when comparing different inferred and simulated distributions of the selection coefficient ($s$)
    is differing conventions in the literature regarding its definition.
    The simulator \slim and the inference tool \grapes define the selection coefficient $s$ such that a homozygote has fitness $1+s$.
    On the other hand, \dadicli and \polydfe define the selection coefficient such that a homozygote has fitness $1+ 2s$.
     %https://academic.oup.com/genetics/article/207/3/1103/5930676
    Moreover, inferred distributions of selection coefficients are typically scaled by an inferred ancestral
    effective population size $N_e$, and different methods assume different scaling.
    \dadicli infers the distribution of $2 N_e s$ while \polydfe and \grapes infer the distribution of $4 N_e s$.
    %https://academic.oup.com/genetics/article/207/3/1103/5930676, https://academic.oup.com/genetics/article/216/2/559/6066183
    For comparison, here we have normalized all inference results to the convention of \slim (see \textbf{Methods}).
    % Add citation to fastDFE as the successor to polyDFE: https://academic.oup.com/mbe/article/41/5/msae070/7641109 ?

@@

    \subsection*{Whole genome simulations}
    In this study we performed simulations of the human and vaquita porpoise genomes
    using the \stdpopsim{} framework.
    All simulations were executed using the \slim engine with a burn-in of $2N$ generations
    (where $N$ is the initial population size), and a scaling factor of 2 (to reduce running time).
    Each simulation was implemented using a relatively short python script, similar to the one shown in Figure 1B
    (see \textbf{Data availablity} below).
    \igcomment{We should add a link to the analysis2 repo in data/code availablity section}
    %https://github.com/popsim-consortium/analysis2/tree/main
    % DRS: Maybe we should say something about how easy it was to do these simulations, and maybe even show
    % some example code somewhere to prove our point? Could help emphasize the utility of this resource.
    % IG: added a note about this above.

    \subsubsection*{Simulations of complete human genomes}
    Simulation of human genomes used a genetic map from the HapMap Project
    \citep{international2007second} (\stdpopsim label \texttt{HapMapII\_GRCh38}) and
    the out of Africa (OOA) demographic model with archaic admixture from
    \cite{ragsdale2019models} (\stdpopsim label \texttt{OutOfAfricaArchaicAdmixture\_5R19}).
    Simulations with selection additionally applied a DFE model with gamma-distributed selection coefficients inferred by
    \cite{kim2017inference} (\stdpopsim label \texttt{Gamma\_K17}) to exons annotated by the HAVANA group release 104
    (\cite{ensembl2018}; \stdpopsim label \texttt{ensembl\_havana\_104\_exons}).
    Each simulation generated 100 diploid genomes from each of the three extant human populations (YRI, CEU, and CHB),
    with each genome consisting of 22 autosomes.
    Using these settings, we generated three replicate datasets without selection
    and three replicate datasets with selection.
    Three additional datasets, used in Figure \ref{fig:homsap-dfe.constant},
    were simulated with the same model of selection and a simple demographic model with one population with a constant size of $10,000$.
    Each of these simulations generated 100 diploid genomes.

@@

    % We used three software packages to infer the distribution of fitness effects (DFE) from genetic data:
    % \dadicli \citep{Huang2023,kim2017inference}, \polydfe \citep{tataru2020polydfe}, and \grapes \citep{galtier2016adaptive}.
    % Each software package was run using the same set of simulations.
    % For these analyses we used the exon annotations from the human and vaquita porpoise genomes respectively.
    % A key distinction between our analyses and those done in practice
    % is that we have complete knowledge of which sites are under selection in the simulation,
    % thus our inferences is a best case scenario. Empirical inference of the DFE is a much
    % harder problem as some mutations assumed to be free of selection
    % may in reality be under selection. As above, all methods were run using default settings.
    \igcomment{I edited the description below written by Silas. Please check to make sure it's still accurate.}
    As described above, each DFE-inference method scales selection coefficients slightly differently.
    We thus rescaled each program's output to obtain a selection coefficient ($s$) consistent with the one used by \slim.
    The output of \grapes  was scaled by a factor of  $\mu / \theta$,
    and the output of \polydfe was saled by $2 \mu / \theta$, 
    where $\mu$ is the assumed mutation rate and 
    $\theta$ is the population-scaled mutation rate estimated by each method ($\theta = 4N_{e}\mu$).
    \igcomment{Do we really multiple $\alpha$ by $\nu$ in rescaling for \dadicli?}
    For \dadicli, we report $\alpha * \nu / (\theta / (4 \mu L p))$, where
    $\alpha$ and $\nu$ are the inferred shape and rate parameters of the \dadicli's inferred gamma distribution,
    $L$ is the total number of sites available for the analysis,
    $p$ is the proportion of sites under selection.

@@

    To test methods for detecting selective sweeps, we simulated 100 evenly distributed 5cM
    segments of human chromosome 1.
    As recombination rate varies across the genome, the size of the region in base pairs simulated varied, however the recombination distance,
    which is critical for the linked effects of a sweep, was held constant.
    Each segment was simulated in each of the following three settings:
    (1) neutral, (2) background selection (BGS), and (3) hard sweep with BGS.
    In all three settings, we applied a genetic map from the HapMap Project
    \citep{international2007second} (\stdpopsim label \texttt{HapMapII\_GRCh38}) and
    the three-population out of Africa (OOA) demographic model from
    \cite{gutenkunst2009inferring} (\stdpopsim label \texttt{OutOfAfrica\_3G09}).
    In the two settings with BGS, we applied the DFE model with gamma-distributed selection coefficients inferred by
    \cite{kim2017inference} (\stdpopsim label \texttt{Gamma\_K17}) to exons annotated by the HAVANA group release 104
    (\cite{ensembl2018}; \stdpopsim label \texttt{ensembl\_havana\_104\_exons}).
    Hard sweeps were modeled by generating a single beneficial mutation 
    with selection coefficient of $s = 0.03$ in the center of the 5cM genomic segment
    and discarding simulations in which the beneficial mutation did not reach (or exceed) a terminal frequency of $95\%$.
    For each of the 100 genomic segments of chromosome 1, we generated 1,000 replicate datasets under each of the three settings.
    As with the whole-genome simulations (see above), we used the \slim engine with a burn-in of $2N$ generations
    (where $N$ is the initial population size), and a scaling factor of 2.
    Each simulation generated 10 diploid genomic segments for each of the two sampled populations: YRI and CEU.
    The sequences simulated with a hard sweep and BGS were used for sweep detection,
    and sequences simulated under the other two settings were used to calibrate the thresholds of the test statistics
    used by the three sweep-detection methods (see below).

@@

\section*{Data availability}\label{data_availability}
The release of \stdpopsim is available at \url{https://github.com/popsim-consortium/stdpopsim}.
This paper coincides with version 0.3.XX of \stdpopsim. %be sure to check this
The entire analysis pipeline associated with this paper is available
as a snakemake workflow at \url{https://github.com/popsim-consortium/analysis2}.
